---
title: "Data Report on Somerset Levels Project"
output: html_notebook
author: "Daniel Smith"
---

```{r message=FALSE, warning=FALSE, include=FALSE}
# Package list:

library(tidyverse)
library(kableExtra)
library(knitr)
library(ggcorrplot)
library(patchwork)
```

```{r include=FALSE}
# Load the processed data for exploration:

load("rdat/01-processing.RData")
```

Abundances of four mosquito species larvae aggregated across all sites and plots for the years 2009-2011 and each season that was recorded. Additionally aggregated abundances are proved for all seasons, as well as a grand total of all abundances across seasons. Species *Cs. morsitans*, *Cs. cantans*, *Cs. caspius* have been removed from the analysis because they were virtually absent in all plots. 2010 is the only complete year where samples were taken throughout all seasons. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gt)

sum1 <- dftidy_long %>% 
  # Grouping vars
  group_by(species, year, season) %>% 
  # Total abundance for all plots and sites
  summarise(abundance = sum(abundance, na.rm = T)) %>%
  # Ungroup
  ungroup() %>% 
  # Change names to nicer ones
  mutate(species = case_when(
    species == "an_claviger" ~ "An. claviger",
    species == "an_maculipennis" ~ "An. maculipennis",
    species == "cs_annulata" ~ "Cs. annulata",
    species == "cx_pipiens" ~ "Cx. pipiens"
  ),
  season = case_when(
    season == 1 ~ "May",
    season == 2 ~ "Jun/Jul",
    season == 3 ~ "Aug/Sep"
  )) %>% 
  # Grouping vars
  group_by(species, year) %>%
  # Aggregate total column
  mutate("All Season" = sum(abundance, na.rm = T)) %>% 
  # spread data
  pivot_wider(names_from = season, values_from = abundance) %>% 
  # Replace NA values with 0
  mutate_if(is.numeric, ~ replace_na(., replace = 0)) %>% 
  # Rename nicely
  rename(Species = species)

# vars names for easier insertion
set1 = colnames(sum1[c(-1, -2)])

# Generic starting table for summarising abundances
tab1 <- sum1 %>% 
  gt(rowname_col = "Species", groupname_col = c("year")) %>% 
  tab_spanner(label = "Season", columns = vars(set1)) %>% 
  tab_stubhead("Species")

# Make formatting and other modifications
tab1 %>% 
  # Move columns around
  cols_move_to_start(columns = "May") %>% 
  cols_move_to_end("All Season") %>% 
  # add a summary row for each year grouping
  summary_rows(
    groups = NULL,
    columns = vars("Jun/Jul", "Aug/Sep", "May", "All Season"),
    fns = list(Total = ~ sum(.)), 
    decimals = 0) %>% 
  
  # this isn't working as intended: totals are wrong
  # # Add a grand total that groups abundances across years for each season
  # grand_summary_rows(
  #   columns = vars(`Jun/Jul`, `Aug/Sep`, `May`, `All Season`),
  #   fns = list(`Grand Total` = ~ sum(),
  #              # Min = ~min(.),
  #              Max = ~max(.),
  #              Mean = ~ mean(.)), 
  #   decimals = 0
  # ) %>% 

# Formatting colour, alignment and size
opt_row_striping(TRUE) %>% 
  opt_align_table_header(align = "center") %>% 
  cols_width(everything() ~ px(175))

```


#### Structural characteristics of Tier 1 and Tier 3 management practices compared:

Waterboard level (or freeboard) is consistently lower in tier 3 management in all seasons. This is expected and means the water is never allowed to drop below 0.35m of the land level in guidance with the tier 3 management practice. Average width of rhynes is slightly larger for tier 3 managed plots across all seasons. Depths at both the centre and proximal edges of rhynes are fairly consistent throughout both management tiers, with slight variations in season 1 and 2. 

The majority of plots were recorded as pasture for tier 1 sites and rough pasture for tier 3 sites. This shows clearly the reduced grazing pressure and reduced management of rhynes in tier 3 plots as depicted by the scheme guidelines. Differences in exposure are slight, with more exposed sunny sights 

```{r echo=FALSE, fig.height=11.7, fig.width=8.3, message=FALSE, warning=FALSE}
# waterbaord level by tier and season
p1 <- dftidy_long %>% 
  group_by(management, season) %>% 
  summarise(
    Width = mean(width, na.rm = T),
    Waterboard = mean(waterboard, na.rm = T)
  ) %>%
  ggplot(aes(x = season, y = Waterboard, fill = management)) +
  geom_col(position = position_dodge()) +
  ylab("Mean Waterboard Level (m)") +
  xlab(NULL) +
  # scale_fill_discrete(name = "Management",
  #                     breaks = c("1", "3"),
  #                     labels = c("Tier 1", "Tier 3"))
  theme(legend.position = "none")

# mean width of rhyne by tier and season
p2 <- dftidy_long %>% 
  group_by(management, season) %>% 
  summarise(
    Width = mean(width, na.rm = T),
    Waterboard = mean(waterboard, na.rm = T)
  ) %>%
  ggplot(aes(x = season, y = Width, fill = management)) +
  geom_col(position = position_dodge()) +
  ylab("Mean Rhyne Width (m)") +
  xlab(NULL) +
  scale_fill_discrete(name = "Management",
                      breaks = c("1", "3"),
                      labels = c("Tier 1", "Tier 3"))

p1.1 <- dftidy_long %>% 
  group_by(management, season) %>% 
  summarise(
    centredepth = mean(centredepth, na.rm = T),
    edgedepth = mean(edgedepth, na.rm = T)
  ) %>%
  ggplot(aes(x = season, y = centredepth, fill = management)) +
  geom_col(position = position_dodge()) +
  # geom_boxplot() +
  ylab("Mean Centre Depth (m)") +
  xlab("Season") +
  scale_fill_discrete(name = "Management",
                      breaks = c("1", "3"),
                      labels = c("Tier 1", "Tier 3"))

p2.1 <- dftidy_long %>% 
  group_by(management, season) %>% 
  summarise(
    centredepth = mean(centredepth, na.rm = T),
    edgedepth = mean(edgedepth, na.rm = T)
  ) %>%
  ggplot(aes(x = season, y = edgedepth, fill = management)) +
  geom_col(position = position_dodge()) +
  # geom_boxplot() +
  ylab("Mean Proximal Depth (m)") +
  xlab("Season") +
  # scale_fill_discrete(name = "Management",
  #                     breaks = c("1", "3"),
  #                     labels = c("Tier 1", "Tier 3"))
  theme(legend.position = "none")



# relative frequency of being Rough or Normal Pasture for Tier 1 and tier 3
p3 <- dftidy_wide %>% 
  filter(landcover == "pasture" | landcover == "rough pasture") %>% 
  count(landcover, management, year) %>% 
  group_by(management) %>% 
  mutate(freq = n / sum(n)) %>% 
  ggplot(aes(x = management, y = freq, fill = landcover)) +
  geom_col() +
  ylab("Frequency") +
  xlab(NULL) +
  scale_fill_discrete(name = "Landcover Type",
                      breaks=c("pasture", "rough pasture"),
                      labels=c("Pasture", "Rough Pasture"))

# Exposure
p4 <- dftidy_long %>%
  count(management, season, exposure) %>% 
  drop_na() %>% 
  group_by(management) %>%
  mutate(freq = n / sum(n)) %>% 
  ggplot(aes(x = management, y = freq, fill = exposure)) +
  geom_col() +
  # Formatting
  xlab(NULL) +
  ylab("Frequency") +
  scale_fill_discrete(name = "Exposure", breaks = c("1", "2", "3", "4", "5"),
        labels = c("Sun", "Partial Shade", "Shade", "Emergent Veg", "Dense Veg"))

# % Shaded by season

p5 <- dftidy_long %>% 
  ggplot(aes(x = management, y = shaded * 100, fill = season)) +
  geom_boxplot() +
  # Formatting 
  xlab("Management Tier") +
  ylab("Shaded (%)") +
  scale_fill_discrete(name = "Season")

# plot with patchwork
(p1 + p2) / (p2.1 + p1.1) / p3 / p4 / p5

```



#### Water chemistry characteristics:
```{r echo=FALSE, fig.height=5.8, fig.width=8.3}
# dftidy_long %>% 
#   ggplot(aes(x = management, y = wtemp, fill = season)) +
#   geom_boxplot()
# 
# dftidy_long %>% 
#   ggplot(aes(x = management, y = turbidity, fill = season)) +
#   geom_boxplot()
# 
# dftidy_long %>% 
#   ggplot(aes(x = management, y = salinity, fill = season)) +
#   geom_boxplot()
# 
# dftidy_long %>% 
#   ggplot(aes(x = management, y = do, fill = season)) +
#   geom_boxplot()
# 
# 
# dftidy_long %>% 
#   ggplot(aes(x = management, y = ph, fill = season)) +
#   geom_boxplot()

chemvar <- c("wtemp", "turbidity", "salinity", "do", "ph")

chemkey <- c(do = "Dissolved Oxygen", ph = "pH", salinity = "Salinity",
             turbidity = "Turbidity", wtemp = "Water Temperature")

pw <- dftidy_long %>% 
  select(site, management, season, plot, chemvar) %>% 
  pivot_longer(all_of(chemvar)) %>% 
  ggplot(aes(x = management, y = value, fill = season)) +
  geom_boxplot() +
  facet_wrap(~ name, scales = "free", labeller = labeller(name = chemkey)) +
  scale_fill_discrete(name = "Season") +
  theme(legend.position = c(.85, .25)) +
  ylab("Value") +
  xlab("Management Tier")

pw

```









